---
title: "Computing segregation indices from existing packages"
author: "C. Cottineau-Mugadza & I. Urria Yañez"

date: today  
format:
  html:
    toc: true
    html-math-method: katex
    
editor: visual
bibliography: assets/References.bib 
csl: assets/apa.csl
link-citations: true
theme: united 
---

## Introduction

Context and aim. Process.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(sf)
```

## Data

```{r dataprep}
#| include: false

Rotterdam <- st_read("data/OECD_fuas/fuas.shp") |>
  filter(fuaname_en == "Rotterdam") |>
  st_transform(28992) |>
  st_buffer(dist = 100)

muni <- st_read("data/gemeenten.gpkg") |>
  st_transform(28992) |>
  st_intersection(Rotterdam) |>
  st_write("processed_data/rotterdam_muni.gpkg")

grid2025 <-  st_read("data/CBS_grids/2025/cbs_vk500_2025_v1.gpkg") |>
  st_transform(28992) |>
  st_intersection(Rotterdam) |>
  mutate(older65 = if_else(aantal_inwoners_65_jaar_en_ouder < 0, NA, aantal_inwoners_65_jaar_en_ouder)) |>
  st_write("processed_data/rotterdam_500m.gpkg")


```

We are using open data from the Dutch statistics office CBS, and in particular 500x500 grid cells of Rotterdam. This is count-data coming from the Dutch register, as the 500m x 500m resolution [@cbs500]. The data was clipped to fit the Rotterdam functional area as defined by the OECD [@oecdFUA]. Municipal boundaries come from the kadaster [@kadaster].

```{r mapRotterdam}

rotterdam_muni <- st_read("processed_data/rotterdam_muni.gpkg") |>
  st_transform(28992) 

rotterdam2025 <-  st_read("processed_data/rotterdam_500m.gpkg") |>
  st_transform(28992) 

FENIX <- rotterdam2025 |> filter(crs28992res500m == "E0925N4350")

ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = older65), linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
  geom_sf(data = FENIX, colour = "yellow", linewidth=0.8) 

```

## Dissimilarity
Ignacio

## Entropy
Clémentine

## Index comparison




