---
title: "Computing segregation indices from existing packages"
author: "C. Cottineau-Mugadza & I. Urria Yañez"

date: today  
format:
  html:
    toc: true
    html-math-method: katex
    
editor: visual
bibliography: assets/References.bib 
csl: assets/apa.csl
link-citations: true
theme: united 
---

## Introduction

Context and aim. Process.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(sf)
```

## Data

```{r dataprep}
#| eval: false
#| message: false
#| warning: false
#| include: false

Rotterdam <- st_read("data/OECD_fuas/fuas.shp") |>
  filter(fuaname_en == "Rotterdam") |>
  st_transform(28992) |>
  st_buffer(dist = 100)

muni <- st_read("data/gemeenten.gpkg") |>
  st_transform(28992) |>
  st_intersection(Rotterdam) |>
  st_write("processed_data/rotterdam_muni.gpkg")

muni <- st_read("data/water.gpkg") |>
  st_transform(28992) |>
  st_intersection(Rotterdam) |>
  st_write("processed_data/rotterdam_water.gpkg")

grid2025 <-  st_read("data/CBS_grids/2025/cbs_vk500_2025_v1.gpkg") |>
  st_transform(28992) |>
  st_intersection(Rotterdam) |>
  mutate(
    across(
      starts_with("aantal_inwoners"),
      ~ if_else(.x < 0, NA, .x)
    )
  ) |>
  rename(
    population = aantal_inwoners,
    age_0_15    = aantal_inwoners_0_tot_15_jaar,
    age_15_25   = aantal_inwoners_15_tot_25_jaar,
    age_25_45   = aantal_inwoners_25_tot_45_jaar,
    age_45_65   = aantal_inwoners_45_tot_65_jaar,
    age_older65 = aantal_inwoners_65_jaar_en_ouder
  ) |>
  mutate(
    age_under65 = age_0_15 + age_15_25 + age_25_45 + age_45_65
  ) |>
  select(crs28992res500m, population, starts_with("age_")) |>
  st_write("processed_data/rotterdam_500m.gpkg")




```

We are using open data from the Dutch statistics office CBS, and in particular 500x500 grid cells of Rotterdam. This is count-data coming from the Dutch register, as the 500m x 500m resolution [@cbs500]. The data was clipped to fit the Rotterdam functional area as defined by the OECD [@oecdFUA]. Municipal boundaries and waterbodies come from the kadaster [@kadaster].

### Population density and spatial distribution

```{r mapRotterdam}
#| message: false
#| warning: false
#| verbose: false


rotterdam_muni <- st_read("processed_data/rotterdam_muni.gpkg") |>
  st_transform(28992) 
rotterdam_water <- st_read("processed_data/rotterdam_water.gpkg") |>
  st_transform(28992) 

rotterdam2025 <-  st_read("processed_data/rotterdam_500m.gpkg") |>
  st_transform(28992) 

FENIX <- rotterdam2025 |> filter(crs28992res500m == "E0925N4350")


ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = population), linewidth=0) +
  geom_sf(data = rotterdam_water, fill = "lightblue", linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
 # geom_sf(data = FENIX, colour = "#00A99D", linewidth=0.7) +
  scale_fill_gradient(low = "white", high = "black", na.value = "#FFE9C9") +
  theme_minimal()

ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = age_older65), linewidth=0) +
  geom_sf(data = rotterdam_water, fill = "lightblue", linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
  scale_fill_gradient(low = "white", high = "#93278F", na.value = "#FFE9C9")+
  theme_minimal()


ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = age_under65), linewidth=0) +
  geom_sf(data = rotterdam_water, fill = "lightblue", linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
  scale_fill_gradient(low = "white", high = "forestgreen", na.value = "#FFE9C9")+
  theme_minimal()
```

### Ratio of populations

```{r mapshares}
#| message: false
#| warning: false
#| verbose: false
rotterdam2025 <- rotterdam2025 |>
  mutate(share_older65 = age_older65 / population)

ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = share_older65), linewidth=0) +
  geom_sf(data = rotterdam_water, fill = "lightblue", linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
  scale_fill_gradient2(midpoint = 0.25, low = "forestgreen", mid = "white", high = "#93278F", na.value = "#FFE9C9")+
  theme_minimal()

```

## Packages

There are currently (at least) 3 packages available to compute the indices of segregation we will use in this session: dissimilarity and entropy.

### `OasisR`

[`OasisR`](https://cran.rstudio.com/web/packages/OasisR/OasisR.pdf) was deprecated last year but seems to function well again. It provides performs most categorical & ordinal measures of segregation, including two-group, one-group, multigroup, and spatially adjusted indices. In addition, it provides resampling methods to test significance. Its core segregation functions generally use a **matrix-oriented structure**: rows represent spatial units and columns represent population groups.

```{r OasisR}
#| message: false
#| warning: false
#| verbose: false


library(OasisR)
```

### `divseg`

[`divseg`](https://christophertkenny.com/divseg/reference/index.html) performs diversity measures and indices proposed by Massey & Denton along the 5 dimensions of segregation [@massey1988dimensions]. It uses a **data-frame and tidyverse-oriented structure**. Group variables are selected from columns, and the package can work directly with `sf` objects.

```{r divseg}
#| message: false
#| warning: false
#| verbose: false


library(divseg)
```

### `segregation`

[`segregation`](https://www.rdocumentation.org/packages/segregation/versions/1.1.0) implements D, exposure and isolation measures, and entropy-based M and H indices. It uses a **long-data structure** in which the population group, spatial unit, and optional frequency or weight are represented explicitly as variables. The package also includes inference, decomposition, and visualisation tools.

```{r segregationpackage}
#| message: false
#| warning: false
#| verbose: false

library(segregation)
```

## Dissimilarity

The Index of Dissimilarity measures the unevenness with which two population groups are distributed across a set of spatial units. For two groups $x$ and $y$ distributed across $n$ units:

$$D = \frac{1}{2}\sum_{i=1}^{n}
\left|
\frac{x_i}{X} - \frac{y_i}{Y}
\right|$$

For the first comparison we distinguish between people aged **0--25** and everyone aged **26 and older**.

### `OasisR`

`OasisR` contains several versions of the dissimilarity index, including the conventional two-group index.

```{r fDissOasisR}
DIDuncan
```

> `DIDuncan()` takes as input a matrix ***x*** in which rows represent spatial units and columns represent population groups. The values should be population counts, rather than population shares. When more than two groups are provided, the function returns the dissimilarity index for every pair of groups. If the total population count is included as a column, it is considered as an additional population group.

```{r DataOasisR}

age_two_count_sf <- rotterdam2025 |>
  mutate(age_under25 = (age_0_15 + age_15_25),
         age_over25 = (age_25_45 + age_45_65 + age_older65)) |>
  select(crs28992res500m, age_under25, age_over25) |>
  drop_na(age_under25, age_over25) |>
  filter(age_under25 + age_over25 > 0)

age_two_count <- age_two_count_sf |> 
  st_drop_geometry()

age_two_count_matrix <- age_two_count |> 
  select(age_under25, age_over25) |>
  as.matrix()

```

Let's calculate the Dissimilarity Index between the first two groups

```{r DissOasisR}
dissim_OasisR <- DIDuncan(age_two_count_matrix)
dissim_OasisR
```

The output is a **pairwise matrix**. With two groups, the value of interest is:

```{r DissOasisRValue}
dissim_OasisR_value <- dissim_OasisR[1, 2]
dissim_OasisR_value
```

### `divseg`

```{r fDissDivseg}
ds_dissim
```

> Group variables are selected from columns, and the package can work directly with sf objects.

Let's calculate the same dissimilarity index with this package:

```{r DissDivseg}
dissim_divseg <- ds_dissim(
  age_two_count, 
  c(age_under25, age_over25))
head(dissim_divseg)
```

The function returns a table with the global value repeated for the rows of the data.

```{r DissDivsegValue}
dissim_divseg_value <- dissim_divseg[[1]]
dissim_divseg_value
```

### `segregation`

```{r fDisssegregation}
dissimilarity
```

> The segregation package requires data in long form (because most segregation data comes in this form), not in the form of contingency tables.

```{r DissimSegregationData}
age_two_count_long <- age_two_count |>
  select(crs28992res500m, age_under25, age_over25) |>
  pivot_longer(!crs28992res500m, names_to = "group", values_to = "n") |>
  filter(n > 0)
```

Let's calculate the dissimilarity index with this package using the long format data:

```{r DissSegregation}
dissim_segregation <- segregation::dissimilarity(
  age_two_count_long,
  group = "group",
  unit = "crs28992res500m",
  weight = "n"
)

dissim_segregation
```

The function returns a one-row table identifying the statistic and its estimate.

```{r DissSegregationValue}
dissim_segregation_value <- dissim_segregation$est[[1]]
dissim_segregation_value
```

### Comparing the same two-group index

All three functions implement the conventional two-group Index of Dissimilarity, even though the required input and returned output are different.

```{r compare-D}
tibble(
  package = c("OasisR", "divseg", "segregation"),
  D = c(
    dissim_OasisR_value,
    dissim_divseg_value,
    dissim_segregation_value
  )
)
```

This is a useful benchmark: if the three implementations represent the same mathematical index and use the same observations, we expect the estimates to agree apart from rounding or implementation details.

**To discuss**: Why using different input and output structures? What is their added value?

### What happens with more than two groups?

The two-group example collapses the full age structure of Rotterdam into a binary distinction. We can instead retain the five age categories available in the CBS data.

```{r MultigroupData}
age_multi_sf <- rotterdam2025 |>
  select(
    crs28992res500m,
    age_0_15,
    age_15_25,
    age_25_45,
    age_45_65,
    age_older65
  ) |>
  drop_na() |>
  filter(
    age_0_15 + age_15_25 + age_25_45 + age_45_65 + age_older65 > 0
  )

age_multi_matrix <- age_multi_sf |>
  st_drop_geometry() |>
  select(-crs28992res500m) |>
  as.matrix()

age_multi_long <- age_multi_sf |>
  st_drop_geometry() |>
  pivot_longer(
    cols = starts_with("age_"),
    names_to = "group",
    values_to = "n"
  )
```

With `OasisR`, `DIDuncan()` returns all pairwise comparisons when more than two groups are supplied:

```{r PairwiseOasisR}
pairwise_D <- DIDuncan(age_multi_matrix)
pairwise_D
```

`OasisR` also provides `DMulti()`, which implements a genuine multigroup generalisation of D:

```{r MultigroupOasisR}
DMulti

dissim_multigroup <- DMulti(age_multi_matrix)
dissim_multigroup
```

The other packages make different design choices. `divseg::ds_dissim()` accepts more than two selected columns, but treats the **first column as the focal group and the sum of the remaining columns as the comparison group**. It therefore remains a two-group calculation.

```{r MultigroupDivseg}
divseg_first_vs_rest <- ds_dissim(
  age_multi_sf,
  c(age_0_15, age_15_25, age_25_45, age_45_65, age_older65),
  .name = "D_under15_vs_rest"
)

first(divseg_first_vs_rest$D_under15_vs_rest)
```

By contrast, `segregation::dissimilarity()` requires the grouping variable to contain exactly two groups. Passing the five-category long data to the function produces an error rather than implicitly redefining the comparison.

```{r MultigroupSegregation}
#| eval: false

segregation::dissimilarity(
  age_multi_long,
  group = "group",
  unit = "crs28992res500m",
  weight = "n"
)
```

**Discuss:** What should a function do when the user supplies more than two groups?

- return all pairwise comparisons?
- collapse the data to one group versus all others?
- calculate a multigroup extension?
- require the user to choose explicitly?

### Adding spatial information

Conventional D depends on the population counts in the spatial units, but not on the locations of those units relative to one another. Rearranging the same grid cells spatially would therefore leave D unchanged.

Morrill's spatial version of D adjusts the conventional index using contiguity between spatial units. `OasisR::DIMorrill()` accepts the group matrix together with a binary contiguity matrix.

We create a rook-contiguity matrix for the same grid cells used in the two-group example. `spdep` is used here only to construct the spatial relationship between cells.

```{r SpatialWeights}
#| message: false
#| warning: false

neighbours <- spdep::poly2nb(
  age_two_count_sf,
  queen = FALSE
)

contiguity <- spdep::nb2mat(
  neighbours,
  style = "B",
  zero.policy = TRUE
)
```

```{r SpatialDissOasisR}
DIMorrill

morrill_matrix <- DIMorrill(
  age_two_count_matrix,
  c = contiguity
)

morrill_D <- morrill_matrix[1, 2]
morrill_D
```

```{r CompareSpatialDiss}
tibble(
  index = c("Duncan's D", "Morrill's spatial D"),
  value = c(dissim_OasisR_value, morrill_D)
)
```

## Entropy

We will use the same three packages to compute entropy indices of segregation.

### `OasisR`

```{r TheilOasisR}
HTheil
```

This function takes as input x:

> an object of class matrix (or which can be coerced to that class), where each column represents the distribution of a group within spatial units. The number of columns should be greater than 1 (at least 2 groups are required). You should not include a column with total population, because this will be interpreted as a group.

```{r entropyOasisR}

age_groups_count_sf <- rotterdam2025 |>
  mutate(age_under25 = (age_0_15 + age_15_25),
         age_25_65 = (age_25_45 + age_45_65)) |>
  select(crs28992res500m, age_under25, age_25_65, age_older65) |>
  drop_na(age_under25, age_25_65, age_older65) |>
  filter(age_under25 + age_25_65 + age_older65 > 0)

age_groups_count <- age_groups_count_sf |> 
  st_drop_geometry()

age_groups_count_matrix <- age_groups_count |> 
  select(age_under25, age_25_65, age_older65) |>
  as.matrix()

entropyOasisR <- HTheil(age_groups_count_matrix)
entropyOasisR

```

### `divseg`

```{r Entropydivseg}
ds_entropy
```

This function takes as input a tibble (which means we can keep the `sf` format). You need to specify the columns with which to compute the measure.

> Must be at least 2 columns. If more than 2, treats first column as first group and sum of other columns as second.

```{r entropyDivseg}
group_datadivseg <- age_groups_count |>
  select(age_under25, age_25_65, age_older65)

entropydivsegUnder25 <- ds_entropy(group_datadivseg, starts_with("age"))[[1]]
entropydivsegUnder25
  
group_datadivseg <- age_groups_count |>
  select(age_25_65, age_under25, age_older65)

entropydivseg25_65 <- ds_entropy(group_datadivseg, starts_with("age"))[[1]]
entropydivseg25_65
  
group_datadivseg <- age_groups_count |>
  select(age_older65, age_under25, age_25_65)

entropydivsegolder65 <- ds_entropy(group_datadivseg, starts_with("age"))[[1]]
entropydivsegolder65
```

### `segregation`

`segregation`: offers standards segregation indices (M, H, D + exposure indices), but also addesses sample bias, decomposition of entropy based indices and visalisation. https://www.rdocumentation.org/packages/segregation/versions/1.1.0

```{r Rsegregationpackage}
#| message: false
#| warning: false
#| verbose: false

library(segregation)


entropy
```

> The segregation package continues to use the same long representation used for D.

https://cran.r-project.org/web/packages/segregation/vignettes/segregation.html

```{r entropysegregationtotal}
age_groups_count_long <- age_groups_count |>
  select(crs28992res500m, age_under25, age_25_65, age_older65) |>
  pivot_longer(!crs28992res500m, names_to = "group", values_to = "n") |>
  filter(n > 0)
  
total <- mutual_total(age_groups_count_long, "group", "crs28992res500m", weight = "n")
total
```

mutual_total() returns two related overall multigroup measures: the Mutual Information Index (M) and Theil's normalised Entropy Index (H).

The package can also return unit-level contributions to segregation:

```{r entropysegregationlocal}

local <- mutual_local(age_groups_count_long, "crs28992res500m", "group", weight = "n", wide = T)
local |> 
  head()
```

This output represents **local contributions by spatial unit**, rather than group-specific entropy scores. The distinction is important when comparing outputs across packages.

### Comparing entropy outputs

For the group-versus-rest calculation, `OasisR::HTheil()` and `divseg::ds_entropy()` provide directly comparable outputs:

| Age category  | Package | Entropy value            |
|---------------|---------|--------------------------|
| Under 25      | OasisR  | `r entropyOasisR[1]`     |
| Under 25      | divseg  | `r entropydivsegUnder25` |
| 25 to 65 y.o. | OasisR  | `r entropyOasisR[2]`     |
| 25 to 65 y.o. | divseg  | `r entropydivseg25_65`   |
| Over 65       | OasisR  | `r entropyOasisR[3]`     |
| Over 65       | divseg  | `r entropydivsegolder65` |

For a genuine overall multigroup H, we can compare `OasisR::HMulti()` with the H returned by `segregation::mutual_total()`:

```{r compare-entropy-multi}
HMulti

entropyOasisR_multi <- HMulti(age_groups_count_matrix)
entropyOasisR_multi

entropysegregation_H <- total |>
  filter(stat == "H") |>
  pull(est)

tibble(
  package = c("OasisR", "segregation"),
  H = c(
    entropyOasisR_multi,
    entropysegregation_H
  )
)
```

`divseg::ds_entropy()` is not included in this second table because its documented behaviour with more than two selected columns is to compare the first group with the sum of the remaining groups, rather than to calculate a multigroup H.
