---
title: "Computing segregation indices from existing packages"
author: "C. Cottineau-Mugadza & I. Urria Yañez"

date: today  
format:
  html:
    toc: true
    html-math-method: katex
    
editor: visual
bibliography: assets/References.bib 
csl: assets/apa.csl
link-citations: true
theme: united 
---

## Introduction

Context and aim. Process.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(sf)
```

## Data

```{r dataprep}
#| eval: false
#| message: false
#| warning: false
#| include: false

Rotterdam <- st_read("data/OECD_fuas/fuas.shp") |>
  filter(fuaname_en == "Rotterdam") |>
  st_transform(28992) |>
  st_buffer(dist = 100)

muni <- st_read("data/gemeenten.gpkg") |>
  st_transform(28992) |>
  st_intersection(Rotterdam) |>
  st_write("processed_data/rotterdam_muni.gpkg")

muni <- st_read("data/water.gpkg") |>
  st_transform(28992) |>
  st_intersection(Rotterdam) |>
  st_write("processed_data/rotterdam_water.gpkg")

grid2025 <-  st_read("data/CBS_grids/2025/cbs_vk500_2025_v1.gpkg") |>
  st_transform(28992) |>
  st_intersection(Rotterdam) |>
  mutate(
    across(
      starts_with("aantal_inwoners"),
      ~ if_else(.x < 0, NA, .x)
    )
  ) |>
  rename(
    population = aantal_inwoners,
    age_0_15    = aantal_inwoners_0_tot_15_jaar,
    age_15_25   = aantal_inwoners_15_tot_25_jaar,
    age_25_45   = aantal_inwoners_25_tot_45_jaar,
    age_45_65   = aantal_inwoners_45_tot_65_jaar,
    age_older65 = aantal_inwoners_65_jaar_en_ouder
  ) |>
  mutate(
    age_under65 = age_0_15 + age_15_25 + age_25_45 + age_45_65
  ) |>
  select(crs28992res500m, population, starts_with("age_")) |>
  st_write("processed_data/rotterdam_500m.gpkg")




```

We are using open data from the Dutch statistics office CBS, and in particular 500x500 grid cells of Rotterdam. This is count-data coming from the Dutch register, as the 500m x 500m resolution [@cbs500]. The data was clipped to fit the Rotterdam functional area as defined by the OECD [@oecdFUA]. Municipal boundaries and waterbodies come from the kadaster [@kadaster].

### Population density and spatial distribution

```{r mapRotterdam}
#| message: false
#| warning: false
#| verbose: false


rotterdam_muni <- st_read("processed_data/rotterdam_muni.gpkg") |>
  st_transform(28992) 
rotterdam_water <- st_read("processed_data/rotterdam_water.gpkg") |>
  st_transform(28992) 

rotterdam2025 <-  st_read("processed_data/rotterdam_500m.gpkg") |>
  st_transform(28992) 

FENIX <- rotterdam2025 |> filter(crs28992res500m == "E0925N4350")


ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = population), linewidth=0) +
  geom_sf(data = rotterdam_water, fill = "lightblue", linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
 # geom_sf(data = FENIX, colour = "#00A99D", linewidth=0.7) +
  scale_fill_gradient(low = "white", high = "black", na.value = "#FFE9C9") +
  theme_minimal()

ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = age_older65), linewidth=0) +
  geom_sf(data = rotterdam_water, fill = "lightblue", linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
  scale_fill_gradient(low = "white", high = "#93278F", na.value = "#FFE9C9")+
  theme_minimal()


ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = age_under65), linewidth=0) +
  geom_sf(data = rotterdam_water, fill = "lightblue", linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
  scale_fill_gradient(low = "white", high = "forestgreen", na.value = "#FFE9C9")+
  theme_minimal()
```

### Ratio of populations

```{r mapshares}
#| message: false
#| warning: false
#| verbose: false
rotterdam2025 <- rotterdam2025 |>
  mutate(share_older65 = age_older65 / population)

ggplot() +
  geom_sf(data = rotterdam2025, aes(fill = share_older65), linewidth=0) +
  geom_sf(data = rotterdam_water, fill = "lightblue", linewidth=0) +
  geom_sf(data = rotterdam_muni, colour = "orange", linewidth=0.4) +
  scale_fill_gradient2(midpoint = 0.25, low = "forestgreen", mid = "white", high = "#93278F", na.value = "#FFE9C9")+
  theme_minimal()

```

## Dissimilarity

Ignacio

## Entropy

There are currently (at least) 3 packages available to compute entropy indices of segregation.

### `OasisR`

`OasisR` was deprecated last year but seems to function well again: performs most categorical & ordinal measures of segregation + resampling methods to test significance https://cran.rstudio.com/web/packages/OasisR/OasisR.pdf

```{r OasisR}
#| message: false
#| warning: false
#| verbose: false


library(OasisR)

HTheil
```
This function takes as input x:

> an object of class matrix (or which can be coerced to that class), where each column represents the distribution of a group within spatial units. The number of columns should be greater than 1 (at least 2 groups are required). You should not include a column with total population, because this will be interpreted as a group.

```{r entropyOasisR}
group_data <- as.data.frame(rotterdam2025) |>
  mutate(share_under25 = (age_0_15 + age_15_25) / population,
         share_25_65 = (age_25_45 + age_45_65) / population ) |>
  select(share_under25, share_25_65, share_older65) |>
  filter(share_under25 > 0 & share_25_65 > 0 & share_older65 > 0)

  entropyOasisR <- HTheil(group_data)
  entropyOasisR

```

### `divseg`

`divseg` performs diversity measures and indices proposed by Massey & Denton along the 5 dimensions of segregation https://christophertkenny.com/divseg/reference/index.html

```{r divseg}
#| message: false
#| warning: false
#| verbose: false


library(divseg)
ds_entropy
```


This function takes as input a tibble (which means we can keep the `sf` format). You need to specify the columns with which to compute the measure. 

> Must be at least 2 columns. If more than 2, treats first column as first group and sum of other columns as second.


```{r entropyDivseg}
group_datadivseg <- rotterdam2025 |>
  mutate(share_under25 = (age_0_15 + age_15_25) / population,
         share_25_65 = (age_25_45 + age_45_65) / population ) |>
  select(share_under25, share_25_65, share_older65) |>
  filter(share_under25 > 0 & share_25_65 > 0 & share_older65 > 0)


  entropydivsegunUnder25 <- ds_entropy(group_datadivseg, starts_with("share"))[[1]]
  entropydivsegunUnder25
  
  group_datadivseg <- group_datadivseg |>
    select(share_25_65, share_older65, share_under25)

  entropydivsegun25_45 <- ds_entropy(group_datadivseg, starts_with("share"))[[1]]
  entropydivsegun25_45
  
   group_datadivseg <- group_datadivseg |>
    select(share_older65, share_25_65, share_under25)

  entropydivsegunolder65 <- ds_entropy(group_datadivseg, starts_with("share"))[[1]]
  entropydivsegunolder65
```




### `segregation`

`segregation`: offers standards segregation indices (M, H, D + exposure indices), but also addesses sample bias, decomposition of entropy based indices and visalisation. https://www.rdocumentation.org/packages/segregation/versions/1.1.0

```{r segregationpackage}
#| message: false
#| warning: false
#| verbose: false

library(segregation)


entropy
```

> The segregation package requires data in long form (because most segregation data comes in this form), not in the form of contingency tables.

https://cran.r-project.org/web/packages/segregation/vignettes/segregation.html


```{r entropysegregation}
long_data <- as.data.frame(rotterdam2025) |>
  mutate(under25 = age_0_15 + age_15_25,
         group25_65 = age_25_45 + age_45_65) |>
select(crs28992res500m, under25, group25_65, age_older65) |>
  pivot_longer(!crs28992res500m,names_to = "group", values_to = "n") |>
  filter(n > 0)
  
  total <- mutual_total(long_data, "group", "crs28992res500m", weight = "n")
  local <- mutual_local(long_data, "crs28992res500m", "group", weight = "n", wide = T)
  total
  local
  
  
```

## Index comparison

We end up with the following measures of group segregation:

| Age category  | Package     | Entropy value             |
|---------------|-------------|---------------------------|
| Under 25      | OasisR      | `r entropyOasisR[1]`      |
| Under 25      | divseg      | `r entropydivsegunUnder25`      |
| Under 25      | segregation      | `r   local[3,2]`      |
| 25 to 25 y.o. | OasisR      | `r entropyOasisR[2]`      |
| 25 to 25 y.o. | divseg      | `r entropydivsegun25_45`      |
| 25 to 25 y.o.      | segregation      | `r   local[2,2]`      |
| Over 65       | OasisR      | `r entropyOasisR[3]`      |
| Over 65       | divseg      | `r entropydivsegunolder65`      |
| Over 65     | segregation      | `r   local[1,2]`      |


